Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-01 docs/manuscript.Rmd
updated

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are two competing models for these variables. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation.",
	display=FALSE
)

@@

fig_nums(
	"cit_mr_comparison_figure", 
	caption="Outcomes were simulated to be causally influenced by exposures, with varying degrees of measurement error applied to both. CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality. The value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality. Rows of graphs represent the sample size used in the simulations. For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality.",
	display=FALSE
)

@@

A common approach for inferring causal relationships between two traits is to use genetic instruments in mediation-based analyses. Here we show that when measurement error is present, widely used tools predicated on this approach, such as the causal inference test (CIT) are susceptible to making erroneous inference of both the existence of causality and the causal direction. Increasing sample size for these tests, rather than reducing bias, can often have the effect of increasing certainty in the wrong answer. We demonstrate that Mendelian randomization (MR) is a technique for causal inference that is relatively robust to measurement error, and here we extend the method with new sensitivity analyses to estimate the direction of causality between correlated variables even when the biology of the instrument is unknown. This extension is less susceptible to measurement error than mediation-based analyses and only requires GWAS summary data. We use it to demonstrate empirically that inferring the direction of causality between correleted DNA methylation and gene expression levels is highly susceptible to differences in measurement error between the two assaying platforms. Other 'omic platforms are likely to be similarly susceptible, necessitating the use of sensitivity analyses to evaluate reliability.

@@

MR can be applied to the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner. Here the SNP, which is assumed to have a direct association with the exposure, is used as a surrogate for the exposure. The causal effect of the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure.

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. An important advantage of mediation-based analysis is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence. Such is not the case for MR, where in order to infer the direction of causality the instrument must be known to influence the exposure directly, associating with the outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e. for which of the two variables is the SNP a valid instrument? `r cfig("dag_figure")`). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. For example, the same FTO variant that is associated with body mass index (BMI) is also associated with C-reactive protein (CRP). It is known that this is because BMI causes CRP [@Timpson2011], but supposing we only had *a priori* knowledge that FTO associated with CRP, and we used it to instrument CRP in an MR analysis of CRP against BMI, we would mistakenly infer that CRP causes BMI. If multiple instruments are available for an hypothesised exposure then this problem can be ameliorated [@DaveySmithHemani2014], however this is not always the case, especially in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known. For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?

MR does however have some important advantages over the mediation-based approaches. First, MR is used as a tool to not just deduce the existance of causality, but to also estimate the causal effect size. Second, the mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Third, MR is substantially more robust to there being measurement error in the exposure variable. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement error and measurement imprecision can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result to measurement error. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.

@@

In these analyses the *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for the incorrect model of $y$ causing $x$. For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:

- If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then the correct model is accepted
- If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then the incorrect model is accepted
- If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then an alternative causal model is accepted
- If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no inference is made

For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value, though we should stress that for real analyses it is not good practice to rely on p-values for making causal inference, nor is it reliable to depend on arbitrary significance thresholds [@Sterne2001].

@@

In what follows we assume no horizontal pleiotropy, such that $g$ only has an effect on one of the variables through the other variable. To infer the direction of causality it is desirable to know which of the variables, $x$ or $y$, is being directly influenced by the instrument $g$. This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Appendix 2), formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980]. It is calculated as

@@

The Steiger test for testing if $x \rightarrow y$ is based on evaluating $\\rho_{gx} > \\rho_{gy}$. However, $\\rho_{gx}$ and $\\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\\rho_{g,x_o} > \\rho_{g,y_o}$. To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we calculate the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also calculate the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2.

@@

Simulations were conducted by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_o$, the outcome $y$, the measured values of the outcome $y_o$ and the instrument $g$. In all models $x$ causes $y$ and $g$ is an instrument for $x$. Each variable was simulated such that:

@@

where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, $\alpha_{mx}$ and $\beta_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $\alpha_{my}$ and $\beta_{my}$ are parameters for non-differential measurement error in the outcome $y$.

@@

$$
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
$$

giving a total of 432 combinations of parameters. Simulations using each of these sets of variables were performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simulated variables. Similar patterns of results were obtained for different values of $cor(g, x)$.

@@

Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_o = f(x)$. For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$, and $\epsilon_{mx}$ represents the imprecision in the measurement of $x$. Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that

@@

where $\beta_x$ is the true effect of the exposure on the outcome. In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$. The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. When measurement error is introduced it can be shown using basic covariance properties (Appendix 1) that

@@

Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$. Therefore, if there is any measurement error that incurs imprecision (i.e. $var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT. Note that linear transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions.

@@

We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure. `r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association. We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure. The performance of MR in detecting an association is unrelated to measurement error in. Measurement error in the outcome does not induce a substantive difference in performance between CIT and MR.

@@

We performed simulations to explore the performance of this approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect. `r cfig("cit_mr_comparison_figure")` shows that when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, the CIT is in general much more fallable to reporting a robust causal association for the wrong direction of causality. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.

>>>>

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three competing models for these variables. The causal inference test (CIT) attempts to delineate between them. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation. c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder. ",
	display=FALSE
)

@@

fig_nums(
	"cit_mr_comparison_figure", 
	caption="a) Outcomes were simulated to be causally influenced by exposures as shown in the graph, with varying degrees of measurement error applied to both. CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality. The value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality. Rows of graphs represent the sample size used in the simulations. For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality. b) As in (a) except the simulated model was non-causal, and a genetic confounder induces an association between $x$ and $y$. MR is unable to identify this model, so any significant associations are deemed to be incorrect. Outcome 3 denotes evidence for the correct model for the CIT method.",
	display=FALSE
)

@@

Inference of the causal relationship between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others. We show here, however, that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer. Here we develop an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction with some advantages. First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; and third, it can be performed using only summary level data from genome-wide association studies. We apply the method to infer the causal direction between DNA methylation and gene expression levels. Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but that such high level conclusions are susceptible to differential measurement error between the platforms. We emphasise that the use of several different methods is important to triangulate reliable conclusions.

@@

MR can be applied to the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner. Here the SNP is used as a surrogate for the exposure. Assuming the SNP associates with the outcome only through the exposure, the causal effect of the exposure on the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure. Though difficult to test empirically, this assumption can be relaxed in various ways when multiple instruments are available for a putative exposure [@Bowden2015; @Bowden2016b] and a number of sensitivity tests are now available to improve reliability [@Hemani2016].

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. An important advantage of mediation-based analysis is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence. The CIT can also distinguish between several putative models of association between the traits and the SNP [@Millstein2009; @Millstein2016]. Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the primary effect of the SNP (i.e. for which of the two variables is the SNP a valid instrument? `r cfig("dag_figure")`). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, such that if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. If multiple instruments are available for an hypothesised exposure, which is increasingly typical for complex traits that are analysed in large GWAS consortia, then techniques can be applied to mitigate these issues [@DaveySmithHemani2014]. But these techniques cannot always be applied in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known. For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?

MR does however have some important advantages over the mediation-based approaches. The mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Additionally, when MR assumptions are satisfied the method is substantially more robust to there being measurement error in the exposure variable. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result of this extension to measurement error. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.

@@

In these analyses the *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the direction of $x$ causing $y$, and for the direction model of $y$ causing $x$. For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:

- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then the $x \rightarrow y$ model is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the $y \rightarrow x$ model is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no inference is made
- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the confounding model is accepted ($x \leftarrow g \rightarrow y$).

For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value, though we stress that for real analyses it is not good practice to rely on p-values for making causal inference, nor is it reliable to depend on arbitrary significance thresholds [@Sterne2001].

@@

The method that we will now describe is able to distinguish between two models, $x \rightarrow y$ or $y \rightarrow	x$. Unlike the CIT framework, this approach cannot infer if the true model is $x \leftarrow g \rightarrow y$. We also assume all genetic effects are additive.

To infer the direction of causality it is desirable to know which of the variables, $x$ or $y$, is being directly influenced by the instrument $g$. This can be achieved by assessing which of the two variables has the biggest absolute correlation with $g$ (Appendix 2), formalised by testing for a difference in the correlations $\rho_{gx}$ and $\rho_{gy}$ using Steiger's Z-test for correlated correlations within a population [@Steiger1980]. It is calculated as

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$. As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds. To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also do the same for the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2.

@@

Simulations were conducted by creating variables of sample size $n$ for the exposure $x$, the measured values of the exposure $x_o$, the outcome $y$, the measured values of the outcome $y_o$ and the instrument $g$. One of two models are simulated, the "causal model" where $x$ causes $y$ and $g$ is an instrument for $x$; or the "non-causal model" where $g$ influences a confounder $u$ which in turn causes both $x$ and $y$. Here $x$ and $y$ are correlated but not causally related. Each variable in the causal model was simulated such that:

@@

where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, $\alpha_{mx}$ and $\beta_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $\alpha_{my}$ and $\beta_{my}$ are parameters for non-differential measurement error in the outcome $y$. Similarly in the non-causal model:

$$
\begin{aligned}
g & \sim Binom(2, 0.5) \\
u & = \alpha_g + \beta_g g + \epsilon_g \\
x & = \alpha_u + \beta_u u + \epsilon_{u_x} \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_u + \beta_u u + \epsilon_{u_y} \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
$$

@@

$$
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(g, u)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
$$

giving a total of 432 combinations of parameters. Simulations using each of these sets of variables were performed 100 times, and the CIT and MR methods were applied to each in order to evaluate the causal association of the simulated variables. Similar patterns of results were obtained for different values of $cor(g, x)$ and $cor(g, u)$.

@@

Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_o = f(x)$. For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that

@@

where $\beta_x$ is the true effect of the exposure on the outcome. In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$. The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that

@@

Thus an observational study will find $cov(g, y_o - \hat{y_o}) = 0$ when the true model is causal only when $D = 1$. Therefore, if there is any measurement error that incurs imprecision (i.e. $var(\epsilon_{mx}) \neq 0$) then there will remain an association between $g$ and $y_o | x$, which is in violation of the the 4th condition of the CIT. Note that scale transformation of $x$ or $y$ without any incurred imprecision is insufficient to lead to a violation of the test statistic assumptions, and henceforth mention of measurement error will relate to imprecision unless otherwise stated.

@@

We performed simulations to compare the performance of MR against CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure. `r cfig("causality_exists_tpr")` shows the true positive rates between the CIT and MR for detecting a causal association. We observe that the CIT has lower power in all cases, with performance declining as measurement imprecision increases in the exposure. When MR assumptions are satisfied, notably that it is known on which of $x$ and $y$ the SNP $g$ has a primary influence, the performance of MR in detecting an association is unrelated to measurement error in the exposure. Measurement error in the outcome does reduce power, but does not induce a substantive difference in performance between CIT and MR.

@@

We performed simulations to explore the performance of this approach in comparison to CIT in terms of the rate at which a strong causal relationship is obtained for the correct direction of causality, and the rate at which a strong causal relationship is obtained where the reported direction of causality is incorrect. Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.

`r cfig("cit_mr_comparison_figure")`a shows that, for the "causal model", when $d < 0$ the MR analysis is indeed liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, the CIT is in general more fallable to reporting a robust causal association for the wrong direction of causality. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality. The CIT, unlike the Steiger test for MR, is able to distinguish the "non-causal model" from the "causal model" (Methods, `r cfig("cit_mr_comparison_figure")`b), but it is evident that measurement error will often lead the CIT to evaluate this confounding model as being causal.
